\textbf{Evaluation of machine learning methods for 16S rRNA gene data}

### 2019-05-16 submission/manuscript.Rmd
update

Begüm D. Topçuoğlu${^1}$, Mack Ruffin${^2}$, Jenna Wiens${^3}$, Patrick D. Schloss^1$\dagger$^

@@

2\. Department of Family Medicine and Community Medicine, Penn State Hershey Medical Center, Hershey, PA

3\. Department of Computer Science and Engineering, University or Michigan, Ann Arbor, MI 49109

@@

To showcase reliable ML pipeline and to shed light on how much ML model selection can affect modeling results, we performed an empirical analysis comparing several different ML models using the same dataset and with a robust ML pipeline. We used a previously published colorectal cancer (CRC) study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences from `r paste(nrow(shared))` patients. We built ML models using fecal 16S rRNA gene sequences to predict patients with normal colons or patients with colonic tumors which are called screen relevant neoplasias (SRN). The study had `r paste(sum(sum(meta$dx=="normal")+sum(meta$Dx_Bin=="Adenoma")))` normal and `r paste(sum(sum(meta$dx=="cancer")+sum(meta$Dx_Bin=="adv Adenoma")))` SRN samples. We established modeling pipelines for L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost. Our ML pipeline utilized held-out test data to evaluate generalization and prediction performance of each ML model. The mean test AUROC varied from `r paste(round(test_all[performance_index[1],3], 3))` (std ± `r paste(round(test_all[performance_index[1],4], 3))`) to `r paste(round(test_all[performance_index[7],3], 3))` (std ± `r paste(round(test_all[performance_index[7],4], 3))`). Random forest had the highest mean AUROC for detecting SRN. Despite the simplicity, the L2-regularized logistic regression followed random forest in performance. In terms of computational efficiency, L2 logistic regression trained the fastest (`r paste(round(walltime_df[walltime_index[1],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[1],3], 3))`), while XGBoost took the longest (`r paste(round(walltime_df[walltime_index[7],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[7],3], 3))`). We found that mean cross-validation and testing AUROC varied only `r paste(round(difference[difference_index[7]], 3))`, which highlights the importance of a seperate held-out test set and consistent preprocessing of the data prior to evaluation. Aside from evaluating generalization and classification performances for each of these models, this study established standards for modeling pipelines of microbiome-associated machine learning models.

@@

__The prediction and generalization performance of seven ML models during cross-validation and when applied to the held-out test data.__

We evaluated the prediction performance of seven binary classification models when applied to held-out test data over 100 data-splits using the area under the receiver operating characteristic curve (AUROC) as the discriminative performance metric. Random forest had the highest mean AUROC for detecting SRN, `r paste(round(mean((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`), followed by L2 logistic regression and XGBoost with mean AUROC values of `r paste(round(mean((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`) and `r paste(round(mean((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="XGBoost"))$test_aucs), 3))`) respectively [Figure 2]. Random forest had significanly higher predictive performances than L2 SVM with linear and radial basis function kernels and decision tree which had mean AUROC values of `r paste(round(mean((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`),  `r paste(round(mean((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`), and `r paste(round(mean((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2]. We also evaluated the generalization performance of each ML model by comparing their mean cross-validation AUROC and mean testing AUROC. We found that mean cross-validation and testing AUROC difference for each model was below 0.01.

@@

\newpage
\includegraphics{Figure_1}
**Figure 1. Machine learning pipeline showing predictive model training and evaluation flowchart.  **  We split the data 80%/20% stratified to maintain the overall label distribution, performed five-fold cross-validation on the training data to select the best hyperparameter setting and then using these hyperparameters to train all of the training data. The model was evaluated on a held-out set of data (not used in selecting the model).
Abbreviations: AUROC, area under the receiver operating characteristic curve
\newpage
\includegraphics{Figure_2.png}
**Figure 2. Generalization and classification performance of ML models using AUROC values of all cross validation and testing performances. **   The median AUROC for diagnosing individuals with SRN using bacterial abundances was higher than chance (depicted by horizontal line at 0.50) for all the ML models. Discriminative performance of random forest model was higher than other ML models. The boxplot shows quartiles at the box ends and the statistical median as the horizontal line in the box. The whiskers show the farthest points that are not outliers. Outliers are data points that are not within 3/2 times the interquartile ranges. 
Abbreviations: SRN, screen-relevant neoplasias; AUROC, area under the receiver operating characteristic curve; SVM, support vector machine; XGBoost, extreme gradient boosting
\newpage
\includegraphics{Figure_3.png}
**Figure 3. Interpretation of the linear ML models.** (A) L2 logistic regression coefficients  (B) L1 SVM with linear kernel feature weights (C) L2 SVM with linear kernel feature weights. The means weights and coefficients of the most important five OTUs for each model are shown here with the standard deviation over 100 data-splits. Similar OTUs had the largest impact on the predictive performance of L2 logistic regression and L2 SVM with linear kernel. 
Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic Unit.
\newpage
\includegraphics{Figure_4.png}
**Figure 4. Explanation of the non-linear ML models.** (A) SVM with radial basis kernel (B) decision tree (C) random forest (D) XGboost feaure importances were explanied using permutation importance using held-out test set. For all the tree-based models, a *Peptostreptococcus* species (OTU00367) had the largest impact on predictive performance of the model. 
Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic Unit; RBF, radial basis kernel; OTU, Operational Taxonomic Unit.
\newpage
\includegraphics{Figure_5.png}
**Figure 5. Computational efficiency of seven ML models.** The walltimes for training and testing of each data-split showed the differences in computational efficieny of the seven models. The median walltime in hours was the highest for XGBoost and shortest for L2 logistic regression. The boxplot shows quartiles at the box ends and the statistical median as the horizontal line in the box. The whiskers show the farthest points that are not outliers. Outliers are data points that are not within 3/2 times the interquartile ranges. 
Abbreviations: AUROC, area under the receiver operating characteristic curve; SVM, support vector machine; XGBoost, extreme gradient boosting.  
\newpage

>>>>

Begüm D. Topçuoğlu${^1}$, Jenna Wiens${^2}$, Mack Ruffin${^3}$, Patrick D. Schloss^1$\dagger$^

@@

2\. Department of Computer Science and Engineering, University or Michigan, Ann Arbor, MI 49109

3\. Department of Family Medicine and Community Medicine, Penn State Hershey Medical Center, Hershey, PA

@@

To showcase a reliable ML pipeline and to shed light on how much ML model selection can affect modeling results, we performed an empirical analysis comparing several different ML models using the same dataset and with a robust ML pipeline. We used a previously published colorectal cancer (CRC) study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences from `r paste(nrow(shared))` patients. We built ML models using fecal 16S rRNA gene sequences to predict patients with normal colons or patients with colonic tumors which are called screen relevant neoplasias (SRN). The study had `r paste(sum(sum(meta$dx=="normal")+sum(meta$Dx_Bin=="Adenoma")))` normal and `r paste(sum(sum(meta$dx=="cancer")+sum(meta$Dx_Bin=="adv Adenoma")))` SRN samples. We established modeling pipelines for L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost. Our ML pipeline utilized held-out test data to evaluate generalization and prediction performance of each ML model. The median test AUROC varied from `r paste(round(test_all[performance_index[1],3], 3))`  to `r paste(round(test_all[performance_index[7],3], 3))`. Random forest had the highest median AUROC for detecting SRN. Despite the simplicity, the L2-regularized logistic regression followed random forest in performance. In terms of computational efficiency, L2 logistic regression trained the fastest (`r paste(round(walltime_df[walltime_index[1],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[1],3], 3))`), while XGBoost took the longest (`r paste(round(walltime_df[walltime_index[7],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[7],3], 3))`). We found that median cross-validation and testing AUROC varied only `r paste(round(abs(difference[difference_index[7]]), 3))`, which highlights the importance of a seperate held-out test set and consistent preprocessing of the data prior to evaluation. Aside from evaluating generalization and classification performances for each of these models, this study established standards for modeling pipelines of microbiome-associated machine learning models.

@@

__Random forest has the highest prediction performance when applied to the held-out test data compared the other six ML models .__

We evaluated the prediction performances of seven binary classification models when applied to held-out test data using the area under the receiver operating characteristic curve (AUROC) as the discriminative performance metric. Random forest had significantly higher test AUROC values than the other models for detecting SRNs. The median AUROC of random forest was `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`). L2 logistic regression, XGBoost, L2 SVM with linear and radial basis function kernel AUROC values were not significantly different from one another. They had median AUROC values of `r paste(round(median((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="XGBoost"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`) and  `r paste(round(median((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`) respectively. L1 SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(round(median((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))`) and `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2].

__Predictive performances of the seven ML models during cross-validation and when applied to the held-out test data do not show large differences.__

We evaluated the generalization performance of each ML model by comparing their median cross-validation AUROC and median testing AUROC. The largest difference between the two was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1 SVM with linear kernel followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively.

@@

\newpage
\includegraphics{Figure_1}
**Figure 1. Machine learning pipeline showing predictive model training and evaluation flowchart.  **  We split the data 80%/20% stratified to maintain the overall label distribution, performed five-fold cross-validation on the training data to select the best hyperparameter setting and then using these hyperparameters to train all of the training data. The model was evaluated on a held-out set of data (not used in selecting the model).
Abbreviations: AUROC, area under the receiver operating characteristic curve
\newpage
\includegraphics{Figure_2.png}
**Figure 2. Generalization and classification performance of ML models using AUROC values of all cross validation and testing performances. **   The median AUROC for diagnosing individuals with SRN using bacterial abundances was higher than chance (depicted by horizontal line at 0.50) for all the ML models. Discriminative performance of random forest model was higher than other ML models. The boxplot shows quartiles at the box ends and the statistical median as the horizontal line in the box. The whiskers show the farthest points that are not outliers. Outliers are data points that are not within 3/2 times the interquartile ranges. 
Abbreviations: SRN, screen-relevant neoplasias; AUROC, area under the receiver operating characteristic curve; SVM, support vector machine; XGBoost, extreme gradient boosting
\newpage
\includegraphics{Figure_3.png}
**Figure 3. Interpretation of the linear ML models.** (A) L2 logistic regression coefficients  (B) L1 SVM with linear kernel feature weights (C) L2 SVM with linear kernel feature weights. The means weights and coefficients of the most important five OTUs for each model are shown here with the standard deviation over 100 data-splits. Similar OTUs had the largest impact on the predictive performance of L2 logistic regression and L2 SVM with linear kernel. 
Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic Unit.
\newpage
\includegraphics{Figure_4.png}
**Figure 4. Explanation of the non-linear ML models.** (A) SVM with radial basis kernel (B) decision tree (C) random forest (D) XGboost feaure importances were explanied using permutation importance using held-out test set. The gray rectangle and the dashed line show the IQR range and median of the base testing AUROC without any permutation performed. For all the tree-based models, a *Peptostreptococcus* species (OTU00367) had the largest impact on predictive performance of the model. 
Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic Unit; RBF, radial basis kernel; OTU, Operational Taxonomic Unit.
\newpage
\includegraphics{Figure_5.png}
**Figure 5. Computational efficiency of seven ML models.** The walltimes for training and testing of each data-split showed the differences in computational efficieny of the seven models. The median walltime in hours was the highest for XGBoost and shortest for L2 logistic regression. The boxplot shows quartiles at the box ends and the statistical median as the horizontal line in the box. The whiskers show the farthest points that are not outliers. Outliers are data points that are not within 3/2 times the interquartile ranges. 
Abbreviations: AUROC, area under the receiver operating characteristic curve; SVM, support vector machine; XGBoost, extreme gradient boosting.  
\newpage

### 2019-05-16 submission/manuscript.tex
update

Begüm D. Topçuoğlu\({^1}\), Mack Ruffin\({^2}\), Jenna Wiens\({^3}\),
Patrick D. Schloss\textsuperscript{1\(\dagger\)}

@@

2. Department of Family Medicine and Community Medicine, Penn State
Hershey Medical Center, Hershey, PA

3. Department of Computer Science and Engineering, University or
Michigan, Ann Arbor, MI 49109

@@

To showcase reliable ML pipeline and to shed light on how much ML model
selection can affect modeling results, we performed an empirical
analysis comparing several different ML models using the same dataset
and with a robust ML pipeline. We used a previously published colorectal
cancer (CRC) study (3) which had fecal 16S rRNA gene sequences from 490
patients. We built ML models using fecal 16S rRNA gene sequences to
predict patients with normal colons or patients with colonic tumors
which are called screen relevant neoplasias (SRN). The study had 261
normal and 229 SRN samples. We established modeling pipelines for
L2-regularized logistic regression, L1 and L2 support vector machines
(SVM) with linear and radial basis function kernels, a decision tree,
random forest and XGBoost. Our ML pipeline utilized held-out test data
to evaluate generalization and prediction performance of each ML model.
The mean test AUROC varied from 0.598 (std ± 0.044) to 0.697 (std ±
0.037). Random forest had the highest mean AUROC for detecting SRN.
Despite the simplicity, the L2-regularized logistic regression followed
random forest in performance. In terms of computational efficiency, L2
logistic regression trained the fastest (0.202 hours, std ± 0.019),
while XGBoost took the longest (162.843 hours, std ± 3.986). We found
that mean cross-validation and testing AUROC varied only 0.01, which
highlights the importance of a seperate held-out test set and consistent
preprocessing of the data prior to evaluation. Aside from evaluating
generalization and classification performances for each of these models,
this study established standards for modeling pipelines of
microbiome-associated machine learning models.

@@

\textbf{The prediction and generalization performance of seven ML models
during cross-validation and when applied to the held-out test data.}

We evaluated the prediction performance of seven binary classification
models when applied to held-out test data over 100 data-splits using the
area under the receiver operating characteristic curve (AUROC) as the
discriminative performance metric. Random forest had the highest mean
AUROC for detecting SRN, 0.697 (std ± 0.037) followed by L2 logistic
regression and XGBoost with mean AUROC values of 0.676 (std ± 0.042) and
0.678 (std ± 0.04) respectively {[}Figure 2{]}. Random forest had
significanly higher predictive performances than those as well as L2
linear SVM, RBF SVM and decision tree which had mean AUROC values of
0.671 (std ± 0.045), 0.663 (std ± 0.044), and 0.598 (std ± 0.044),
respectively {[}Figure 2{]}. We also evaluated the generalization
performance of each ML model by comparing their mean cross-validation
AUROC and mean testing AUROC. We found that mean cross-validation and
testing AUROC difference for each model was below 0.01.

@@

The ML models we built using L2-regularized logistic regression, L1 and
L2 support vector machines (SVM) with linear and radial basis function
kernels, a decision tree, random forest and XGBoost increase in
complexity as they decrase in interpretibility, respectively. We
interpreted L1 and L2 SVM with linear kernel using the feature weights
and L2 logistic regression using regression coefficients of the trained
models. We calculated the mean weights and coefficients of all the
features over the 100 data-splits. In the three linear models, OTUs that
had the largest mean weights and drove the detection of SRNs belong to
family \emph{Lachnospiraceae}, and \emph{Ruminococcaceae} (OTU01239,
OTU00659, OTU00742, OTU00012, OTU00050, OTU00015, OTU00768, OTU00822,
OTU00609, OTU01212, OTU00629) and genera \emph{Gamella} (OTU00426)
{[}Figure 3{]}. We explained the feature importances in non-linear
models using permutation importance on the held-out test data. For the
tree-based models, permuting \emph{Peptostreptococcus} (OTU00367)
abundances randomly, dropped the predictive performances the most.
Decision tree, random forest and XGBoost models' predictive performance
dropped from 0.6 base testing AUROC median to 0.52, from 0.69 to 0.68
and from 0.68 to 0.65, respectively {[}Figure 4{]}. The negative
predictive impact of \emph{Peptostreptococcus} in the decision tree
model was followed by a \emph{Lachnospiraceae} species (OTU00058) (0.6
base testing AUROC median to 0.58) {[}Figure 4B{]}. Other OTUs had none
to minimal effect on the predictive performance.

@@

The data used for this analysis are stool bacterial abundances and
clinical information of the patients recruited by Great Lakes-New
England Early Detection Research Network study. These data were obtained
from Sze et al (15). The stool samples were provided by recruited adult
participants who were undergoing scheduled screening or surveillance
colonoscopy. Colonoscopies were performed and fecal samples were
collected from participants in four locations: Toronto (ON, Canada),
Boston (MA, USA), Houston (TX, USA), and Ann Arbor (MI, USA). Patients'
colonic disease status was defined by colonoscopy with adequate
preparation and tissue histopathology of all resected lesions. Patients
with an adenoma greater than 1 cm, more than three adenomas of any size,
or an adenoma with villous histology were classified as advanced
adenoma. Study had 172 patients with normal colonoscopies, 198 with
adenomas and 120 with carcinomas. Of the 198 adenomas, 109 were
identified as advanced adenomas. Stool provided by the patients was used
for 16S rRNA gene sequencing to measure bacterial population abundances.
The bacterial abundance data was generated by Sze et al, by processing
16S rRNA sequences in Mothur (v1.39.3) using the default quality
filtering methods, identifying and removing chimeric sequences using
VSEARCH and assigning to OTUs at 97\% similarity using the OptiClust
algorithm (16--18).

@@

\includegraphics{Figure_1} \textbf{Figure 1. Machine learning pipeline
showing predictive model training and evaluation flowchart. } We split
the data 80\%/20\% stratified to maintain the overall label
distribution, performed five-fold cross-validation on the training data
to select the best hyperparameter setting and then using these
hyperparameters to train all of the training data. The model was
evaluated on a held-out set of data (not used in selecting the model).
Abbreviations: AUROC, area under the receiver operating characteristic
curve \newpage
\includegraphics{Figure_2.png} \textbf{Figure 2. Generalization and
classification performance of ML models using AUROC values of all cross
validation and testing performances. } The median AUROC for diagnosing
individuals with SRN using bacterial abundances was higher than chance
(depicted by horizontal line at 0.50) for all the ML models.
Discriminative performance of random forest model was higher than other
ML models. The boxplot shows quartiles at the box ends and the
statistical median as the horizontal line in the box. The whiskers show
the farthest points that are not outliers. Outliers are data points that
are not within 3/2 times the interquartile ranges. Abbreviations: SRN,
screen-relevant neoplasias; AUROC, area under the receiver operating
characteristic curve; SVM, support vector machine; XGBoost, extreme
gradient boosting \newpage
\includegraphics{Figure_3.png} \textbf{Figure 3. Interpretation of the
linear ML models.} (A) L2 logistic regression coefficients (B) L1 SVM
with linear kernel feature weights (C) L2 SVM with linear kernel feature
weights. The means weights and coefficients of the most important five
OTUs for each model are shown here with the standard deviation over 100
data-splits. Similar OTUs had the largest impact on the predictive
performance of L2 logistic regression and L2 SVM with linear kernel.
Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic
Unit. \newpage
\includegraphics{Figure_4.png} \textbf{Figure 4. Explanation of the
non-linear ML models.} (A) SVM with radial basis kernel (B) decision
tree (C) random forest (D) XGboost feaure importances were explanied
using permutation importance using held-out test set. For all the
tree-based models, a \emph{Peptostreptococcus} species (OTU00367) had
the largest impact on predictive performance of the model.
Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic
Unit; RBF, radial basis kernel; OTU, Operational Taxonomic Unit.
\newpage
\includegraphics{Figure_5.png} \textbf{Figure 5. Computational
efficiency of seven ML models.} The walltimes for training and testing
of each data-split showed the differences in computational efficieny of
the seven models. The median walltime in hours was the highest for
XGBoost and shortest for L2 logistic regression. The boxplot shows
quartiles at the box ends and the statistical median as the horizontal
line in the box. The whiskers show the farthest points that are not
outliers. Outliers are data points that are not within 3/2 times the
interquartile ranges. Abbreviations: AUROC, area under the receiver
operating characteristic curve; SVM, support vector machine; XGBoost,
extreme gradient boosting.\\
\newpage

>>>>

Begüm D. Topçuoğlu\({^1}\), Jenna Wiens\({^2}\), Mack Ruffin\({^3}\),
Patrick D. Schloss\textsuperscript{1\(\dagger\)}

@@

2. Department of Computer Science and Engineering, University or
Michigan, Ann Arbor, MI 49109

3. Department of Family Medicine and Community Medicine, Penn State
Hershey Medical Center, Hershey, PA

@@

To showcase reliable ML pipeline and to shed light on how much ML model
selection can affect modeling results, we performed an empirical
analysis comparing several different ML models using the same dataset
and with a robust ML pipeline. We used a previously published colorectal
cancer (CRC) study (3) which had fecal 16S rRNA gene sequences from 490
patients. We built ML models using fecal 16S rRNA gene sequences to
predict patients with normal colons or patients with colonic tumors
which are called screen relevant neoplasias (SRN). The study had 261
normal and 229 SRN samples. We established modeling pipelines for
L2-regularized logistic regression, L1 and L2 support vector machines
(SVM) with linear and radial basis function kernels, a decision tree,
random forest and XGBoost. Our ML pipeline utilized held-out test data
to evaluate generalization and prediction performance of each ML model.
The median test AUROC varied from 0.601 to 0.695. Random forest had the
highest median AUROC for detecting SRN. Despite the simplicity, the
L2-regularized logistic regression followed random forest in
performance. In terms of computational efficiency, L2 logistic
regression trained the fastest (0.202 hours, std ± 0.019), while XGBoost
took the longest (162.843 hours, std ± 3.986). We found that median
cross-validation and testing AUROC varied only 0.021, which highlights
the importance of a seperate held-out test set and consistent
preprocessing of the data prior to evaluation. Aside from evaluating
generalization and classification performances for each of these models,
this study established standards for modeling pipelines of
microbiome-associated machine learning models.

@@

\textbf{Random forest has the highest prediction performance when
applied to the held-out test data compared the other six ML models .}

We evaluated the prediction performances of seven binary classification
models when applied to held-out test data using the area under the
receiver operating characteristic curve (AUROC) as the discriminative
performance metric. Random forest had significantly higher test AUROC
values than the other models for detecting SRNs. The median AUROC of
random forest was 0.695 (IQR 0.044). L2 logistic regression, XGBoost, L2
SVM with linear and radial basis function kernel AUROC values were not
significantly different from one another. They had median AUROC values
of 0.68 (IQR 0.055), 0.679 (IQR 0.053), 0.678 (IQR 0.056) and 0.668 (IQR
0.056) respectively. L1 SVM with linear kernel and decision tree had
significantly lower AUROC values than the other ML models with median
AUROC of 0.65 (IQR 0.066) and 0.601 (IQR 0.059), respectively {[}Figure
2{]}.

\textbf{Predictive performances of the seven ML models during
cross-validation and when applied to the held-out test data do not show
large differences.}

We evaluated the generalization performance of each ML model by
comparing their median cross-validation AUROC and median testing AUROC.
The largest difference between the two was 0.021 in L1 SVM with linear
kernel followed by SVM with radial basis function kernel and decision
tree with a difference of 0.007 and 0.006, respectively.

@@

The ML models we built using L2-regularized logistic regression, L1 and
L2 support vector machines (SVM) with linear and radial basis function
kernels, a decision tree, random forest and XGBoost increase in
complexity as they decrase in interpretibility. We interpreted L1 and L2
SVM with linear kernel using the feature weights and L2 logistic
regression using regression coefficients of the trained models. We
calculated the mean weights and coefficients of all the features over
the 100 data-splits. In the three linear models, OTUs that had the
largest mean weights and drove the detection of SRNs belong to family
\emph{Lachnospiraceae}, and \emph{Ruminococcaceae} (OTU01239, OTU00659,
OTU00742, OTU00012, OTU00050, OTU00015, OTU00768, OTU00822, OTU00609,
OTU01212, OTU00629) and genera \emph{Gamella} (OTU00426) {[}Figure 3{]}.
We explained the feature importances in non-linear models using
permutation importance on the held-out test data. For the tree-based
models, permuting \emph{Peptostreptococcus} (OTU00367) abundances
randomly, dropped the predictive performances the most. Decision tree,
random forest and XGBoost models' predictive performance dropped from
0.6 base testing AUROC median to 0.52, from 0.69 to 0.68 and from 0.68
to 0.65, respectively {[}Figure 4{]}. The negative predictive impact of
\emph{Peptostreptococcus} in the decision tree model was followed by a
\emph{Lachnospiraceae} species (OTU00058) (0.6 base testing AUROC median
to 0.58) {[}Figure 4B{]}. Other OTUs had none to minimal effect on the
predictive performance.

@@

The data used for this analysis are stool bacterial abundances and
clinical information of the patients recruited by Great Lakes-New
England Early Detection Research Network study. These data were obtained
from Sze et al (15). The stool samples were provided by recruited adult
participants who were undergoing scheduled screening or surveillance
colonoscopy. Colonoscopies were performed and fecal samples were
collected from participants in four locations: Toronto (ON, Canada),
Boston (MA, USA), Houston (TX, USA), and Ann Arbor (MI, USA). Patients'
colonic health was labeled by colonoscopy with adequate preparation and
tissue histopathology of all resected lesions. Patients with an adenoma
greater than 1 cm, more than three adenomas of any size, or an adenoma
with villous histology were classified as advanced adenoma. Study had
172 patients with normal colonoscopies, 198 with adenomas and 120 with
carcinomas. Of the 198 adenomas, 109 were identified as advanced
adenomas. Stool provided by the patients was used for 16S rRNA gene
sequencing to measure bacterial population abundances. The bacterial
abundance data was generated by Sze et al, by processing 16S rRNA
sequences in Mothur (v1.39.3) using the default quality filtering
methods, identifying and removing chimeric sequences using VSEARCH and
assigning to OTUs at 97\% similarity using the OptiClust algorithm
(16--18).

@@

\includegraphics{Figure_1} \textbf{Figure 1. Machine learning pipeline
showing predictive model training and evaluation flowchart. } We split
the data 80\%/20\% stratified to maintain the overall label
distribution, performed five-fold cross-validation on the training data
to select the best hyperparameter setting and then using these
hyperparameters to train all of the training data. The model was
evaluated on a held-out set of data (not used in selecting the model).
Abbreviations: AUROC, area under the receiver operating characteristic
curve \newpage
\includegraphics{Figure_2.png} \textbf{Figure 2. Generalization and
classification performance of ML models using AUROC values of all cross
validation and testing performances. } The median AUROC for diagnosing
individuals with SRN using bacterial abundances was higher than chance
(depicted by horizontal line at 0.50) for all the ML models.
Discriminative performance of random forest model was higher than other
ML models. The boxplot shows quartiles at the box ends and the
statistical median as the horizontal line in the box. The whiskers show
the farthest points that are not outliers. Outliers are data points that
are not within 3/2 times the interquartile ranges. Abbreviations: SRN,
screen-relevant neoplasias; AUROC, area under the receiver operating
characteristic curve; SVM, support vector machine; XGBoost, extreme
gradient boosting \newpage
\includegraphics{Figure_3.png} \textbf{Figure 3. Interpretation of the
linear ML models.} (A) L2 logistic regression coefficients (B) L1 SVM
with linear kernel feature weights (C) L2 SVM with linear kernel feature
weights. The means weights and coefficients of the most important five
OTUs for each model are shown here with the standard deviation over 100
data-splits. Similar OTUs had the largest impact on the predictive
performance of L2 logistic regression and L2 SVM with linear kernel.
Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic
Unit. \newpage
\includegraphics{Figure_4.png} \textbf{Figure 4. Explanation of the
non-linear ML models.} (A) SVM with radial basis kernel (B) decision
tree (C) random forest (D) XGboost feaure importances were explanied
using permutation importance using held-out test set. The gray rectangle
and the dashed line show the IQR range and median of the base testing
AUROC without any permutation performed. For all the tree-based models,
a \emph{Peptostreptococcus} species (OTU00367) had the largest impact on
predictive performance of the model. Abbreviations: SVM, support vector
machine; OTU, Operational Taxonomic Unit; RBF, radial basis kernel; OTU,
Operational Taxonomic Unit. \newpage
\includegraphics{Figure_5.png} \textbf{Figure 5. Computational
efficiency of seven ML models.} The walltimes for training and testing
of each data-split showed the differences in computational efficieny of
the seven models. The median walltime in hours was the highest for
XGBoost and shortest for L2 logistic regression. The boxplot shows
quartiles at the box ends and the statistical median as the horizontal
line in the box. The whiskers show the farthest points that are not
outliers. Outliers are data points that are not within 3/2 times the
interquartile ranges. Abbreviations: AUROC, area under the receiver
operating characteristic curve; SVM, support vector machine; XGBoost,
extreme gradient boosting.\\
\newpage
